---
title: "Your Paper Title Here"
author: "Meredith Carter, Maggie Lam, and Swarnya Srivastava"
date: 3/6/2026
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(LearnBayes)
library(ggplot2)
library(dplyr)
library(janitor)
library(fastR2)
library(HDInterval)
```

# Abstract
Write your abstract here.

# Keywords
*Bayesian Statistics*, *Likelihood Inference*, *Relative Risk*, *Efficacy*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df), ylab = "COVID", 
           main = "COVID Status based on Vaccine vs Placebo")
```

# Statistical Methods

## Model
Here we introduce the statistical model used for inference in this paper. Denote the random variable $T$ as the number in the BNT162b2 vaccine group from the $n=170$ COVID-19 cases. The distribution of $T$ can be well approximated by a binomial distribution (see Appendix proof A for proof). Then $$T\sim Binom(n=170, \pi)$$  where $\pi$ denotes the probability of an individual being in the vaccine group given they contracted COVID-19.

Let $n_1$, $\pi_v$ and $n_2$, $\pi_p$ be the number of individuals and probability an individual contracts COVID-19 in the vaccine and placebo groups respectively. $\pi=P(\text{vaccine}|\text{COVID-19})$ and by Bayes' theorem can be shown to be equal to  $$\pi=\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p}$$

We observe $n_1=17411$ and $n_2=17511$, and since $n_1\approx n_2$, the randomization is 1:1 and we simplify $\pi$ to $$\pi=\frac{\pi_v}{\pi_v+\pi_p}$$

We also observe that our event rates $\pi_v$ and $\pi_p$ are small, so we are will use the ratio between the probabilities, or the relative risk $RR = \frac{\pi_v}{\pi_p}$. The estimated relative risk for the COVID-19 vaccine trials is $\widehat{RR} = \frac{\widehat{\pi}_v}{\widehat{\pi}_p} = \frac{0.0004593}{0.0004593} \approx 0.05010$. This means that the risk of contracting COVID-19 in the vaccinated group is around 5\% that of the placebo group. This also means that the vaccine reduces the risk of contracting COVID-19 by around 95\%, which is the vaccine efficacy.

Denote the parameter of interest $\psi$ to be the vaccine efficacy defined as $1-RR$:
$$\psi = 1-\frac{\pi_v}{\pi_p} = \frac{\pi_p-\pi_a}{\pi_p}$$
Writing this in terms of the binomial probability $\pi$, we have our parameter of interest for inference:
$$\psi=\frac{1-2\pi}{1-\pi}$$

Using this single binomial random variable model, we will calculate estimators for $\psi_0$, the true efficacy of the vaccine, and make inference within two frameworks: Likelihood and Bayes. We will calculate 95% confidence interval estimates for $\psi_0$ and conduct a significance test at the $\alpha=0.05$ level for testing $$H_0:\psi_0=0.3 \text{ vs. } H_1:\psi_0>0.3$$

Altogether, $T$ denotes the number from the vaccinated group among the 170 COVID-19 cases. We assume that $T \sim Binom(170, \pi)$, and we observe $t=8$. 

## Likelihood Inference
Detail the likelihood approach.

We know that the likelihood function for $\pi$ will be $L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162}$, which means the likelihood function of $\psi$ would be $L^*(\psi) = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162}$ or $\ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi))$. From this, we can see that $\hat\psi^{mle}_0 = 0.9506173$, which would be a global maximum. 

### Confidence Interval
We will construct 95\% confidence intervals for $\psi_0$ in 2 ways: using a large sample confidence interval estimate and a parametric bootstrap percentile interval.

#### Large sample confidence interval estimate
For our Binomial model $T$, we assume 2 regularity conditions. First, we assume that the possible values of $t$ do not depend on our parameter $\psi$, and second, that the true value of the parameter does not fall on a boundary. These conditions can be verified by plotting the log-likelihood function of $\psi$ along with the second-order Taylor approximation around the MLE and checking if the second-order Taylor series well approximates our log-likelihood function.

```{r second_order_approximation, cache=TRUE, warning=FALSE, echo=FALSE}
loglik_binom <- function(pi, n, x) {
  ifelse((pi <= 0) | (pi >= 1), NA,
    log(choose(n, x)) + x * log(pi) + (n - x) * log(1 - pi)
  )
}

ml <- maxLik2(loglik = loglik_binom, start = 0.5, n = 170, x = 8)

plot(ml) +
  labs(title = "Second Order Approximation to Binomial Log-Likelihood",
       x = expression(pi))
```

Based on the figure, we can assume these regularity conditions hold. And so we can say $$\hat\psi_0^{mle}\approx Norm\left(\psi_0, \sqrt{\frac{1}{I(\psi_0)}}\right)$$ where $I(\psi_0)$ is the expected Fisher information defined as (see appendix B):

$$I(\psi_0)=nE\left[-\frac{d^2}{d\psi^2}\ln(f_\psi(X))\right]_{\psi=\psi_0}=\frac{n}{(1-\psi)(2-\psi)^2}$$

Then we can then calculate a $100(1-0.05)\%$ large sample confidence interval for $\psi_0$ with the formula: 
$$\hat\psi_0^{mle} \pm z_{\alpha/2}\sqrt{\frac{1}{I(\hat\psi_0^{mle})}}$$

#### Parametric Bootstrap Percentile Interval

With this approach, we will assume samples denoted as $T^{*}$ are drawn from the binomial distribution $T\sim Binom(170, \hat\pi^{mle})$. For each generated sample, we can calculate a bootstrapped estimate of $\hat\pi$, $\hat\pi^{*}=\frac{T^{*}}{n}$. Then, calculate a bootstrap estimate of $\psi$ using the known relationship between $\psi$ and $\pi$, $$\hat\psi^*=\frac{1-2\hat\pi^{*}}{1-\hat\pi^{*}}=\frac{1-2\frac{T^{*}}{n}}{1-\frac{T^{*}}{n}}$$

We will generate 1000 boostrap estimates of $\hat\psi$, and find the 2.5th and 97.5th percentiles of the distribution of these generated estimates to get a $100(1-0.05)\%$ confidence interval for $\psi_0$.


### Significance Test
We wish to test $$H_0: \psi_0=0.3, H_1: \psi_0 > 0.3$$
We can use the likelihood ratio statistic to calculate a p-value: $$\Lambda = \frac{L\left(\widehat{\psi}_0^{mle}\right)}{L\left(\psi_0^{null}\right)}$$
When the null hypothesis $H_0$ is true, $2\ln\Lambda \approx \chi^2_1$, when $n \rightarrow \infty$. The approximate $\chi^2_1$ distribution on which the likelihood ratio significance test is based relies on the second order approximation to the log-likelihood being accurate. We can verify this by observing the graph of the log likelihood function with its quadratic approximation as seen in figure [LABEL].

Our log-likelihood function doesn't closely resemble its quadratic approximation and isn't symmetric around the MLE, which is a reason to concerned about approximating our likelihood ratio with a $\chi_1^2$ distribution. We can verify our results from the p-value from the $\chi^2_1$ distribution by calculating an empirical p-value.

We know that $L(\psi) = {170 \choose 8} (\frac{1 - \psi}{2 - \psi})^8(1 - \frac{1 - \psi}{2 - \psi})^{162}$, so we get the following values for $L\left(\widehat{\psi}_0^{mle}\right)$ and $L\left(\psi_0^{null}\right)$:

\begin{align*}
\widehat{\psi}_0^{mle} &= 0.9506173\\
L\left(\widehat{\psi}_0^{mle}\right) &= L(0.9506173)\\
&= {170 \choose 8} \left(\frac{1 - 0.9506173}{2 - 0.9506173}\right)^8\left(1 - \frac{1 - 0.9506173}{2 - 0.9506173}\right)^{162}\\
&= 0.142988121\\
\psi_0^{null} &= 0.3\\
L\left(\psi_0^{null}\right) &= L(0.3) \\
&= {170 \choose 8} \left(\frac{1 - 0.3}{2 - 0.3}\right)^8\left(1 - \frac{1 - 0.3}{2 - 0.3}\right)^{162}\\
&= 5.62253317 \times 10^{-28}
\end{align*}

We use the values above to calculate the likelihood ratio statistic $\Lambda$.

\begin{align*}
\Lambda &= \frac{L\left(\widehat{\psi}_0^{mle}\right)}{L\left(\psi_0^{null}\right)}\\
&= \frac{0.142988121}{5.62253317 \times 10^{-28}}\\
&= 2.543126 \times 10^{-26}\\
W &= 2\ln \Lambda\\
&= 2\ln \left( 2.543126 \times 10^{-26} \right)\\
&= 121.6012
\end{align*}

We now calculate the p-value $P(\chi^2_1 \geq W)=P(\chi^2_1 \geq 121.6012)$.

```{r significance_test, echo=FALSE, eval = F}
p_value <- pchisq(121.6012, df=1, lower.tail=FALSE)
p_value
```

With a p-value of $2.822313 \times 10^{-28}$, we reject the null hypothesis $H_0: \psi_0=0.3$ under an alpha level $\alpha=0.05$, supporting our alternative hypothesis $H_1: \psi_0 > 0.3$ that the vaccine efficacy is greater than 0.3.

An alternative to using the $\chi^2_1$ distribution to calculate the p-value is to calculate the empirical p-value by generating a large number of samples from the null hypothesis. We then determine the fraction of samples with a likelihood ratio test statistic greater than $W$ to calculate the p-value.

```{r empirical_p_value, cache=TRUE, echo=FALSE}
set.seed(414)
n <- 170
x <- 8
psi_null <- 0.3
pi_null <- (1 - psi_null) / (2 - psi_null)
pi_mle <- summary(ml)$estimate[1]
W_obs <- 2 * ((log(choose(n, x)) + x * log(pi_mle) + (n - x) * log(1 - pi_mle)) -
              (log(choose(n, x)) + x * log(pi_null) + (n - x) * log(1 - pi_null)))

B <- 10000
w_star <- numeric(B)

for (i in 1:B) {
  x_star <- rbinom(1, size = n, prob = pi_null)
  pi_mle_star <- x_star/n
  
  w_star[i] <- 2 * ((log(choose(n, x_star)) + 
                       x_star * log(pi_mle_star) + 
                       (n - x_star) * log(1 - pi_mle_star)) -
              (log(choose(n, x_star)) + 
                 x_star * log(pi_null) + (n - x_star) * log(1 - pi_null)))
}

hist(w_star, breaks = 20, main = "Distribution of LRT statistic", xlab = expression(W))

p_value_empirical <- sum(w_star > W_obs)/B
p_value_empirical
```
With an empirical p-value of 0, we reach the same conclusion as we did with the p_value we reached assuming the likelihood ration test statistic $W$ follows a $\chi^2_1$ distribution. We reject the null hypothesis $H_0: \psi_0=0.3$ under an alpha level $\alpha=0.01$, supporting our alternative hypothesis $H_1: \psi_0 > 0.3$ that the vaccine efficacy is greater than 0.3.

## Bayesian Inference
Detail the Bayesian approach.

For our inference, we decided to use the values $P(\psi \geq 0.3) = .01 \Rightarrow P(\pi \leq 7/17) = .05$ and $P(\psi \geq 0) = .1 \Rightarrow P(\pi \leq 1/2) = .1$. We chose these values, since we wanted to be very pessimistic with our initial assumption, so we are absolutely certain that the vaccine works. By doing this, it means that the data must be extremely different from our prior assumptions in order for the vaccine to be proven to be more than 30\% effective, so the FDA will know without a doubt that the vaccine is effective in preventing COVID when compared to the placebo. 

```{r, meredith_bayes_ab, echo = F}
ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))
```

From this, we can see that we would have an a value of `r ab[1]` and a b value of `r ab[2]`. This would give us the following models: $$f(T | \pi) \sim Binom(n = 170, \pi) \text{ Data Model}$$ $$g(\pi) = Beta(a = `r ab[1]`, b = `r ab[2]`) \text{ Prior Model}$$ $$h(\pi | t) = Beta(a = `r ab[1]` - t, b = 170 - t + `r ab[2]`)$$

```{r, meredith_bayes_vis, echo = F}
lik_binom <- function(pi, n, x){
  10^2 * dbinom(x = x, size = n, prob = pi)
}

ggplot() + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  stat_function(fun = lik_binom,
                mapping = aes(color = "Likelihood"),
                args = list(n = 170, x = 8),
                xlim = c(0,1), n = 500) +
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(pi), 
       y = "PDF")
```

### Confidence Interval

```{r summarize_pi_posterior_dist, echo=FALSE, eval = F}
posterior_median_pi <- qbeta(0.5, shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
cat("posterior median ", round(posterior_median_pi, 4), "\n")
# equal tail 95% interval
qbeta(p = c(0.025, 0.975), shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
# 95% HPDI for pi (if asymmetric)
hdi(qbeta, credMass = 0.95, shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
```
Our Bayesian model gives us a posterior median of 0.0728 for $\pi$. This means $P(\pi<0.0728\:|\:t)=0.5$. Assuming our posterior distribution of $\pi$ is symmetric, our Bayesian model gives us a 95\% confidence interval for $\pi$ of $(0.0416, 0.1154)$. This means $P(0.0416 < \pi < 0.1154\:|\:t) = 0.95$. Alternatively, the narrowest possible 95\% confidence interval for $\pi$ is $(0.0391, 0.1120)$. This also means $P(0.0391 < \pi < 0.1120\:|\:t) = 0.95$. Since $\psi = \frac{1-2\pi}{1-\pi} \Rightarrow \pi=\frac{1-\psi}{2-\psi}$, we can write 
\begin{align*}
P(\pi<0.0728\:|\:t)&=0.5\\
P\left(\frac{1-\psi}{2-\psi}<0.0728\:|\:t \right)&=0.5\\
P(1-\psi<0.0728(2-\psi)|t)&=0.5\\
P\left(\psi>\frac{1-2\cdot 0.0728}{1-0.0728}|t\right)&=0.5\\
P(\psi>0.9215)&=0.5\\\\
P(0.0416 < \pi < 0.1154\:|\:t) &= 0.95\\
P\left(0.0416 < \frac{1-\psi}{2-\psi} < 0.1154\:|\:t \right) &= 0.95\\
P\left(\frac{1-2\cdot 0.1154}{1-0.1154} < \psi < \frac{1-2\cdot 0.0416}{1-0.0416} \:|\:t\right) &= 0.95\\
P(0.8695 < \psi < 0.9566) &= 0.95\\\\
P(0.0391 < \pi < 0.1120\:|\:t) &= 0.95\\
P(0.0391 < \frac{1-\psi}{2-\psi} < 0.1120\:|\:t) &= 0.95\\
P\left(\frac{1-2\cdot 0.1120}{1-0.1120} < \psi < \frac{1-2\cdot 0.0391}{1-0.0391} \:|\:t\right) &= 0.95\\
P(0.8739 < \psi < 0.9593) &= 0.95
\end{align*}
The posterior median of $\psi$ is 0.9215. Assuming our posterior distribution of $\pi$ is symmetric, our Bayesian model gives us a 95\% confidence interval of $(0.8695, 0.9566)$. Alternatively, the narrowest possible 95\% confidence interval for $\pi$ is $(0.8739, 0.9593)$. 

### Significance Test

We will conduct a significance test at the $\alpha=0.05$ level for testing $$H_0:\psi_0=0.3 \text{ vs. } H_1:\psi_0>0.3$$
Since $\psi = \frac{1-2\pi}{1-\pi} \Rightarrow \pi=\frac{1-\psi}{2-\psi}$, we can write our null and alternative hypotheses for testing as 
$H_0: \psi_0=\frac{1-2\pi_0}{1-\pi_0}=0.3 \Rightarrow \pi_0=\frac{1-0.3}{2-0.3}=\frac{7}{17}$ vs. $H_1: \psi_0=\frac{1-2\pi_0}{1-\pi_0}>0.3 \Rightarrow \pi_0<\frac{1-0.3}{2-0.3}=\frac{7}{17}$.

```{r bayes_sig_test, echo=FALSE}
pbeta(q = (7/17), shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2], lower.tail=FALSE)
```
With a p-value of $1.227958 \times 10^{-26}$, we can reject the null hypothesis $H_0: \pi_0=0.4118$ under an alpha level $\alpha=0.05$, supporting our alternative hypothesis $H_1: \pi_0 < 0.4118$. In terms of $\psi$, this corresponds to an alternative hypothesis $H_1: \psi_0 > 0.3$, suggesting the vaccine efficacy is greater than 30\%.

# Results
<!-- TODO: move likelihood + bayesian significance test results here -->
### Likelihood Inference

#### Confidence Intervals
With the large sample estimate confidence interval, we have our expected Fisher Information for $\hat\psi^{mle}$: $$I(\hat\psi_0^{mle})=\frac{n}{(1-\hat\psi^{mle})(2-\hat\psi^{mle})^2}$$
which we can then plug in our observed value of $n=170$ and calculated value of $\hat\psi^{mle}=0.9506173$.

Then our 95\% confidence interval is $$0.9506173 \pm 1.96\sqrt{\frac{1}{I(\hat\psi_0^{mle})}}$$ where 1.96 is the number that cuts off an area of $0.05/2$ to the right under the standard normal curve. So our large sample confidence interval for $\psi_0$ is (see code section in appendix): $$[0.9155621, 0.9856725]$$

With the parametric bootstrap method, we use our estimated $\hat\pi=\frac{8}{170}$ and simulate 1000 samples of $T$ from the binomial distribution. We can see the bootstrapped sampling distribution of $\pi_{hat}$, and then transform to find percentiles to give us a 95\% confidence interval for $\psi_0$ (see mle_bootstrap code in appendix):
$$[0.9102564, 0.9820359]$$

# Discussion / Conclusion

Through our research, we discovered that the BNT162b2 vaccine reduces the likelihood of getting COVID-19 when compared to the placebo by 95\%, which is reinforced through the 2020 Pfizer study's matching result. This means that the vaccine seems to be effective in reducing COVID-19, which the following results affirm. We decided to use two different methods in order to test the efficacy of the vaccine - namely likelihood inference and Bayesian inference Through likelihood inference, we discovered that with the data given, the most likely probability of the efficacy of the vaccine when compared to the placebo is also 95\%, (with a 95\% confidence interval of approximately 91-92\% to approximately 98-99\%) which reaffirms our prior conclusion. Using this method, we calculated a p-value of $2.8*10^{-9}$, meaning we would reject the null hypothesis and conclude that our vaccine is more than 30\% effective and should be approved by the FDA. Similarly, when using Bayesian inference, we used extremely pessimistic prior assumptions in order to prove without a doubt that the vaccine is successful if it does in fact work. Through this method, we found that the most likely effectiveness of the vaccine is 92\%. In addition to this, we found a 95\% confidence interval of 87\% to 96\%, which means we would reject our null hypothesis with a p-value of $1.2*10^{-26}$. From these results, we can see that our conclusions match those of the Pfizer study, and both recommend using the vaccine to prevent the spread of COVID-19, due to its efficacy of over 30\%. 

# Bibliography

Grove, R. (2026). *Maximum Likelihood Estimation, Likelihood Inference, Bayes Prior/Posterior, Bayes Point/Inferval Estimation*. Lecture slides, STAT 342, University of Washington

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Fernando P. Polack, M.D., et al. (2020). *Safety and Efficacy of the BNT162b2 mRNA Covid-19 Vaccine*. Retrieved from https://www.nejm.org/doi/full/10.1056/nejmoa2034577.

# Appendix

## Appendix A: Likelihood Equation

$\pi = \frac{1 - \psi}{2 - \psi} \\ L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162} \\ L(\psi) = {170 \choose 8} (\frac{1 - \psi}{2 - \psi})^8(1 - \frac{1 - \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{2 - \psi - 1 + \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162} \\ \ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi)) \\ \frac{d}{d\psi}\ell(\psi) = \frac{-8}{1 - \psi} + \frac{8}{2 - \psi} + \frac{162}{2 - \psi} = \frac{170}{2 - \psi} - \frac{8}{1 - \psi} = 0 \\ 170 - 170 \psi - 16 + 8\psi = 0 \\ 162\psi = 154 \\ \hat\psi^{mle}_0 = 0.9506173$  

## Appendix B: Expected Fisher Information
\begin{gather*}
\frac{d}{d\psi}\ell(\psi) = \frac{n}{2 - \psi} - \frac{T}{1 - \psi}\\
\frac{d}{d^2\psi}\ell(\psi)= \frac{n}{(2 - \psi)^2} - \frac{T}{(1 - \psi)^2}
\end{gather*}
Then finding $I(\psi)$:
\begin{align*}
  I(\psi_0)&=nE\left[-\frac{d^2}{d\psi^2}\ln(f_\psi(X))\right]_{\psi=\psi_0}\\
  &= E\left[-\frac{d}{d^2\psi}\ell(\psi)\right]_{\psi=\psi_0}\\
  &= E\left[\frac{-n}{(2 - \psi)^2} + \frac{T}{(1 - \psi)^2}\right]_{\psi=\psi_0}\\
   &= E\left[\frac{-n}{(2 - \psi_0)^2} + \frac{T}{(1 - \psi_0)^2}\right]\\
   &= E\left[\frac{T}{(1 - \psi_0)^2}\right] - \frac{n}{(2 - \psi_0)^2} \\
   &= \frac{n\pi}{(1 - \psi_0)^2}- \frac{n}{(2 - \psi_0)^2} \quad&&\text{Expected value of binom}\\
   &= \frac{n\left(\frac{1-\psi}{2-\psi}\right)}{(1 - \psi_0)^2} - \frac{n}{(2 - \psi_0)^2} \\
   &= \frac{n}{(1-\psi)(2-\psi)^2}
\end{align*}


## Code
```{r ref.label = "important_R_code", eval = TRUE}
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df))

loglik_binom <- function(pi, n, x) {
  ifelse((pi <= 0) | (pi >= 1), NA,
    log(choose(n, x)) + x * log(pi) + (n - x) * log(1 - pi)
  )
}

ml <- maxLik2(loglik = loglik_binom, start = 0.5, n = 170, x = 8)

plot(ml) +
  labs(title = "Second Order Approximation to Binomial Log-Likelihood",
       x = expression(pi))

p_value <- pchisq(121.6012, df=1, lower.tail=FALSE)
p_value

set.seed(414)
n <- 170
x <- 8
psi_null <- 0.3
pi_null <- (1 - psi_null) / (2 - psi_null)
pi_mle <- summary(ml)$estimate[1]
W_obs <- 2 * ((log(choose(n, x)) + x * log(pi_mle) + (n - x) * log(1 - pi_mle)) -
              (log(choose(n, x)) + x * log(pi_null) + (n - x) * log(1 - pi_null)))

B <- 10000
w_star <- numeric(B)

for (i in 1:B) {
  x_star <- rbinom(1, size = n, prob = pi_null)
  pi_mle_star <- x_star/n
  
  w_star[i] <- 2 * ((log(choose(n, x_star)) + 
                       x_star * log(pi_mle_star) + 
                       (n - x_star) * log(1 - pi_mle_star)) -
              (log(choose(n, x_star)) + 
                 x_star * log(pi_null) + (n - x_star) * log(1 - pi_null)))
}

hist(w_star, breaks = 20, main = "Distribution of LRT statistic", xlab = expression(W))

p_value_empirical <- sum(w_star > W_obs)/B
p_value_empirical

ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))

lik_binom <- function(pi, n, x){
  10^2 * dbinom(x = x, size = n, prob = pi)
}

ggplot() + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  stat_function(fun = lik_binom,
                mapping = aes(color = "Likelihood"),
                args = list(n = 170, x = 8),
                xlim = c(0,1), n = 500) +
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(pi), 
       y = "PDF")
posterior_median_pi <- qbeta(0.5, shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
cat("posterior median ", round(posterior_median_pi, 4), "\n")
# equal tail 95% interval
qbeta(p = c(0.025, 0.975), shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
# 95% HPDI for pi (if asymmetric)
hdi(qbeta, credMass = 0.95, shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2])
```

``` {r ref.label = "mle_ci", eval = TRUE}
psi_hat <- 0.9506173
z <- 1.96
n <- 170

i <- n / ((1-psi_hat) * (2-psi_hat)^2)
lower <- psi_hat - z * sqrt(1/i)
upper <- psi_hat + z * sqrt(1/i)

lower
upper
```

```{r label="mle_bootstrap", eval=TRUE}
B = 10000
set.seed(414)

boot_df <- tibble(
  pi_hat = replicate(n = B,
                     expr = rbinom(n = 1, size = 170, prob = 8/170) / 170)
)

ggplot(data = boot_df,
       mapping = aes(x=pi_hat)) +
  geom_histogram(binwidth = 0.01) +
  labs (title = "Bootstrapped sampling distribution of ",
        subtitle=expression(hat(pi)[0]))


lower <- quantile(boot_df$pi_hat, 0.05)
upper <- quantile(boot_df$pi_hat, 0.975)

psi_lower <- (1-2*upper[[1]]) / (1-upper[[1]])
psi_upper <- (1-2*lower[[1]]) / (1-lower[[1]])

psi_lower
psi_upper
```

## Proofs
### Proof A
Let $X$ denote the number of COVID-19 cases among the $n_1$ patients randomly assigned to the vaccine group and $Y$ denote the number of COVID-19 cases among the $n_2$ patients randomly assigned to the placebo group. Restricting our attention to the total of 170 COVID-19 cases observed in the study, we want to know the PMF of the random variable $T$ that is defined as the count in the vaccine group given this restriction, or $T=X|X+Y=170$.

Since the sample sizes $n_1$ and $n_2$ are large, and the rate that an individual contracts COVID-19 is small, we can assume $$X\approx Pois(n_1\pi_v)\quad\text{ and }\quad Y\approx Pois(n_2\pi_p)$$ 
where $X$ and $Y$ are independent, since the Poisson distribution approximates a binomial with a large $n$ and small $p$ well, as in this scenario. 

Therefore the PMF of $T$ can be found as follows:
\begin{align*}
P(T=t)&=P(X=t|X_Y=n)\quad\quad\quad && t=0,1,\dots,n\\
&=\frac{P(X=t\cap X+Y=n)}{P(X+Y=n)}\\
&=\frac{P(X=t\cap Y=n-t)}{P(X+Y=n)}\\
&=\frac{P(X=t)\cdot P(Y=n-t)}{P(X+Y=n)}\quad\quad &&\text{$X$ and $Y$ are independent}\\
&=\frac{\frac{e^{(-n_1\pi_v)}{(n_1\pi_v)}^t}{t!} \cdot P(Y=n-t)}{P(X+Y=n)}\\
&=\frac{\frac{e^{(-n_1\pi_v)}{(n_1\pi_v)}^t}{t!} \cdot \frac{e^{(-n_2\pi_p)}{(n_2\pi_p)}^{(n-t)}}{(n-t)!}}{P(X+Y=n)}\\
&=\frac{\frac{e^{(-n_1\pi_v)}{(n_1\pi_v)}^t}{t!} \cdot \frac{e^{(-n_2\pi_p)}{(n_2\pi_p)}^{(n-t)}}{(n-t)!}}{\frac{e^{(-(n_2\pi_p+n_1\pi_v))}{(n_2\pi_p+n_1\pi_v)}^{n}}{n!}}\\
&=\frac{e^{(-n_1\pi_v)}{(n_1\pi_v)}^t}{t!} \cdot \frac{e^{(-n_2\pi_p)}{(n_2\pi_p)}^{(n-t)}}{(n-t)!} \cdot  \frac{n!}{e^{(-(n_2\pi_p+n_1\pi_v))}{(n_2\pi_p+n_1\pi_v)}^{n}}\\
&=\frac{(n_1\pi_v)^t\cdot (n_2\pi_p)^{(n-t)}}{t!(n-t)!} \cdot \frac{n!}{(n_2\pi_p+n_1\pi_v)^n}\\
&=\frac{n!}{t!(n-t)!}\cdot \frac{(n_1\pi_v)^t\cdot (n_2\pi_p)^{(n-t)}}{(n_2\pi_p+n_1\pi_v)^n}\\
&=\binom{n}{t} \frac{(n_1\pi_v)^t\cdot (n_2\pi_p)^{(n-t)}}{(n_2\pi_p+n_1\pi_v)^n}\\
&=\binom{n}{t} \frac{(n_1\pi_v)^t\cdot (n_2\pi_p)^{(n-t)}}{(n_2\pi_p+n_1\pi_v)^t(n_2\pi_p+n_1\pi_v)^{n-t}}\\
&=\binom{n}{t} \left(\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \right)^t \left(\frac{n_2\pi_p}{n_1\pi_v+n_2\pi_p} \right)^{n-t}\\
&=\binom{n}{t} \left(\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \right)^t \left(\frac{n_1\pi_v-n_1\pi_v+n_2\pi_p}{n_1\pi_v+n_2\pi_p} \right)^{n-t}\\
&=\binom{n}{t} \left(\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \right)^t \left(1-\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \right)^{n-t}
\end{align*}
This is precisely the PMF for a Binomial distribution. Therefore, we have shown that 
$$T\sim Binom(n, \pi=\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p})$$

