---
title: "Your Paper Title Here"
author: "Meredith Carter, Maggie Lam, and Swarnya Srivastava"
date: 3/6/2026
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(LearnBayes)
library(ggplot2)
library(dplyr)
library(janitor)
library(fastR2)
```

# Abstract
Write your abstract here.

# Keywords
*Bayesian Statistics*, *Likelihood Inference*, *Relative Risk*, *Efficacy*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df), ylab = "COVID", 
           main = "COVID Status based on Vaccine vs Placebo")
```

# Statistical Methods

## Model
Here we introduce the statistical model used for inference in this paper. Denote the random variable $T$ as the number in the BNT162b2 vaccine group from the $n=170$ COVID-19 cases. The distribution of $T$ can be well approximated by a binomial distribution (see Appendix proof A for proof). Then $$T\sim Binom(n=170, \pi)$$  where $\pi$ denotes the probability of an individual being in the vaccine group given they contracted COVID-19.

Let $n_1$, $\pi_v$ and $n_2$, $\pi_p$ be the number of individuals and probability an individual contracts COVID-19 in the vaccine and placebo groups respectively. $\pi=P(\text{vaccine}|\text{COVID-19})$ and by Bayes' theorem can be shown to be equal to  $$\pi=\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p}$$

We observe $n_1=17411$ and $n_2=17511$, and since $n_1\approx n_2$, the randomization is 1:1 and we simplify $\pi$ to $$\pi=\frac{\pi_v}{\pi_v+\pi_p}$$

We also observe that our event rates $\pi_v$ and $\pi_p$ are small, so we are will use the ratio between the probabilities, or the relative risk $RR = \frac{\pi_v}{\pi_p}$. The estimated relative risk for the COVID-19 vaccine trials is $\widehat{RR} = \frac{\widehat{\pi}_v}{\widehat{\pi}_p} = \frac{0.0004593}{0.0004593} \approx 0.05010$. This means that the risk of contracting COVID-19 in the vaccinated group is around 5\% that of the placebo group. This also means that the vaccine reduces the risk of contracting COVID-19 by around 95\%, which is the vaccine efficacy.

Denote the parameter of interest $\psi$ to be the vaccine efficacy defined as $1-RR$:
$$\psi = 1-\frac{\pi_v}{\pi_p} = \frac{\pi_p-\pi_a}{\pi_p}$$
Writing this in terms of the binomial probability $\pi$, we have our parameter of interest for inference:
$$\psi=\frac{1-2\pi}{1-\pi}$$

Using this single binomial random variable model, we will calculate estimators for $\psi_0$, the true efficacy of the vaccine, and make inference within two frameworks: Likelihood and Bayes. We will calculate 95% confidence interval estimates for $\psi_0$ and conduct a significance test at the $\alpha=0.01$ level for testing $$H_0:\psi_0=0.2 \text{vs. } H_1:\psi_0>0.2$$

Let $T$ be the number from the vaccinated group among the 170 COVID-19 cases. We assume that $T \sim Binom(170, \pi)$ and we observed $t=8$. 

## Likelihood Inference
Detail the likelihood approach.

We know that the likelihood function for $\pi$ will be $L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162}$, which means the likelihood function of $\psi$ would be $L^*(\psi) = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162}$ or $\ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi))$. From this, we can see that $\hat\psi^{mle}_0 = 0.9506173$, which would be a global maximum. 

### Significance Test
We wish to test $$H_0: \psi_0=0.3, H_1: \psi_0 > 0.3$$.
We can use the likelihood ratio statistic to calculate a p-value: $$\Lambda = \frac{L\left(\widehat{\psi}_0^{mle}\right)}{L\left(\psi_0^{null}\right)}$$.
When the null hypothesis $H_0$ is true, $2\ln\Lambda \approx \chi^2_1$, when $n \rightarrow \infty$. The approximate $\chi^2_1$ distribution on which the likelihood ratio significance test is based relies on the second order approximation to the log-likelihood being accurate. We can verify this by observing a graph of the log likelihood function with its quadratic approximation.

```{r second_order_approximation, cache=TRUE, warning=FALSE, echo=FALSE}
loglik_binom <- function(pi, n, x) {
  ifelse((pi <= 0) | (pi >= 1), NA,
    log(choose(n, x)) + x * log(pi) + (n - x) * log(1 - pi)
  )
}

ml <- maxLik2(loglik = loglik_binom, start = 0.5, n = 170, x = 8)

plot(ml) +
  labs(title = "Second Order Approximation to Binomial Log-Likelihood",
       x = expression(pi))
```

Our log-likelihood function doesn't closely resemble its quadratic approximation and isn't symmetric around the MLE, which is a reason to concerned about approximating our likelihood ratio with a $\chi_1^2$ distribution. We can verify our results from the p-value from the $\chi^2_1$ distribution by calculating an empirical p-value.

We know that $L(\psi) = {170 \choose 8} (\frac{1 - \psi}{2 - \psi})^8(1 - \frac{1 - \psi}{2 - \psi})^{162}$, so we get the following values for $L\left(\widehat{\psi}_0^{mle}\right)$ and $L\left(\psi_0^{null}\right)$:

\begin{align*}
\widehat{\psi}_0^{mle} &= 0.9506173\\
L\left(\widehat{\psi}_0^{mle}\right) &= L(0.9506173)\\
&= {170 \choose 8} (\frac{1 - 0.9506173}{2 - 0.9506173})^8(1 - \frac{1 - 0.9506173}{2 - 0.9506173})^{162}\\
&= 0.142988121\\
\psi_0^{null} &= 0.3\\
L\left(\psi_0^{null}\right) &= L(0.3) \\
&= {170 \choose 8} (\frac{1 - 0.3}{2 - 0.3})^8(1 - \frac{1 - 0.3}{2 - 0.3})^{162}\\
&= 5.62253317 \times 10^{-28}
\end{align*}

We use the values above to calculate the likelihood ratio statistic $\Lambda$.

\begin{align*}
\Lambda &= \frac{L\left(\widehat{\psi}_0^{mle}\right)}{L\left(\psi_0^{null}\right)}\\
&= \frac{0.142988121}{5.62253317 \times 10^{-28}}\\
&= 2.543126 \times 10^{-26}\\
W &= 2\ln \Lambda\\
&= 2\ln \left( 2.543126 \times 10^{-26} \right)\\
&= 121.6012
\end{align*}

We now calculate the p-value $P(\chi^2_1 \geq W)=P(\chi^2_1 \geq 121.6012)$.

```{r significance_test, echo=FALSE}
p_value <- pchisq(121.6012, df=1, lower.tail=FALSE)
p_value
```

With a p-value of $2.822313 \times 10^{-28}$, we reject the null hypothesis $H_0: \psi_0=0.3$ under an alpha level $\alpha=0.01$, supporting our alternative hypothesis $H_1: \psi_0 > 0.3$ that the vaccine efficacy is greater than 0.3.

An alternative to using the $\chi^2_1$ distribution to calculate the p-value is to calculate the empirical p-value by generating a large number of samples from the null hypothesis. We then determine the fraction of samples with a likelihood ratio test statistic greater than $W$ to calculate the p-value.

```{r empirical_p_value, cache=TRUE, echo=FALSE}
set.seed(414)
n <- 170
x <- 8
psi_null <- 0.3
pi_null <- (1 - psi_null) / (2 - psi_null)
pi_mle <- summary(ml)$estimate[1]
W_obs <- 2 * ((log(choose(n, x)) + x * log(pi_mle) + (n - x) * log(1 - pi_mle)) -
              (log(choose(n, x)) + x * log(pi_null) + (n - x) * log(1 - pi_null)))

B <- 10000
w_star <- numeric(B)

for (i in 1:B) {
  x_star <- rbinom(1, size = n, prob = pi_null)
  pi_mle_star <- x_star/n
  
  w_star[i] <- 2 * ((log(choose(n, x_star)) + 
                       x_star * log(pi_mle_star) + 
                       (n - x_star) * log(1 - pi_mle_star)) -
              (log(choose(n, x_star)) + 
                 x_star * log(pi_null) + (n - x_star) * log(1 - pi_null)))
}

hist(w_star, breaks = 20, main = "Distribution of LRT statistic", xlab = expression(W))

p_value_empirical <- sum(w_star > W_obs)/B
p_value_empirical
```
With an empirical p-value of 0, we reach the same conclusion as we did with the p_value we reached assuming the likelihood ration test statistic $W$ follows a $\chi^2_1$ distribution. We reject the null hypothesis $H_0: \psi_0=0.3$ under an alpha level $\alpha=0.01$, supporting our alternative hypothesis $H_1: \psi_0 > 0.3$ that the vaccine efficacy is greater than 0.3.

## Bayesian Inference
Detail the Bayesian approach.

For our inference, we decided to use the values $P(\psi \geq 0.3) = .01 \Rightarrow P(\pi \leq 7/17) = .05$ and $P(\psi \geq 0) = .1 \Rightarrow P(\pi \leq 1/2) = .1$. We chose these values, since we wanted to be very pessimistic with our initial assumption, so we are absolutely certain that the vaccine works. By doing this, it means that the data must be extremely different from our prior assumptions in order for the vaccine to be proven to be more than 30\% effective, so the FDA will know without a doubt that the vaccine is effective in preventing COVID when compared to the placebo. 

```{r, meredith_bayes_ab, echo = F}
ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))
```

From this, we can see that we would have an a value of `r ab[1]` and a b value of `r ab[2]`. This would give us the following models: $$f(T | \pi) \sim Binom(n = 170, \pi) \text{ Data Model}$$ $$g(\pi) = Beta(a = `r ab[1]`, b = `r ab[2]`) \text{ Prior Model}$$ $$h(\pi | t) = Beta(a = `r ab[1]` - t, b = 170 - t + `r ab[2]`)$$

```{r, meredith_bayes_vis}
lik_binom <- function(pi, n, x){
  10^2 * dbinom(x = x, size = n, prob = pi)
}

ggplot() + # Icky for some reason, but I think this is what we were supposed to do?
  geom_function(fun = dbeta, # I will look at this again later and see if I can find the error, because something is really wrong
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  stat_function(fun = lik_binom,
                mapping = aes(color = "Likelihood"),
                args = list(n = 170, x = 8),
                xlim = c(0,1), n = 500) +
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(pi), 
       y = "PDF")
```

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Likelihood Equation

$\pi = \frac{1 - \psi}{2 - \psi} \\ L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162} \\ L(\psi) = {170 \choose 8} (\frac{1 - \psi}{2 - \psi})^8(1 - \frac{1 - \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{2 - \psi - 1 + \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162} \\ \ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi)) \\ \frac{d}{d\psi}\ell(\psi) = \frac{-8}{1 - \psi} + \frac{8}{2 - \psi} + \frac{162}{2 - \psi} = \frac{170}{2 - \psi} - \frac{8}{1 - \psi} = 0 \\ 170 - 170 \psi - 16 + 8\psi = 0 \\ 162\psi = 154 \\ \hat\psi^{mle}_0 = 0.9506173$  

## Code
```{r ref.label = "important_R_code", eval=FALSE}
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df))

ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))

lik_binom <- function(pi, n, x){
  10^2 * dbinom(x = x, size = n, prob = pi)
}

ggplot() + # Icky for some reason, but I think this is what we were supposed to do?
  geom_function(fun = dbeta, # I will look at this again later and see if I can find the error, because something is really wrong
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  stat_function(fun = lik_binom,
                mapping = aes(color = "Likelihood"),
                args = list(n = 170, x = 8),
                xlim = c(0,1), n = 500) +
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(pi), 
       y = "PDF")
```

## Proofs
### Proof A
Let $X$ denote the number of COVID-19 cases among the $n_1$ patients randomly assigned to the vaccine group and $Y$ denote the number of COVID-19 cases among the $n_2$ patients randomly assigned to the placebo group. Restricting our attention to the total of 170 COVID-19 cases observed in the study, we want to know the PMF of the random variable $T$ that is defined as the count in the vaccine group given this restriction, or $T=X|X+Y=170$.

Since the sample sizes $n_1$ and $n_2$ are large, and the rate that an individual contracts COVID-19 is small, we can assume $$X\approx Pois(n_1\pi_v)\quad\text{ and }\quad Y\approx Pois(n_2\pi_p)$$ 
where $X$ and $Y$ are independent, since the Poisson distribution approximates a binomial with a large $n$ and small $p$ well, as in this scenario. 

Therefore the PMF of $T$ can be found as follows:
\begin{align*}
P(T=t)&=P(X=t|X_Y=n)\quad\quad\quad && t=0,1,\dots,n\\
&=\frac{P(X=t\cap X+Y=n)}{P(X+Y=n)}\\
&=\frac{P(X=t\cap Y=n-t)}{P(X+Y=n)}\\
&=\frac{P(X=t)\cdot P(Y=n-t)}{P(X+Y=n)}\quad\quad &&\text{Since $X$ and $Y$ are independent}
\end{align*}

