---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(LearnBayes)
```

# Abstract
Write your abstract here.

# Keywords
*Bayesian Statistics*, *Likelihood Inference*, *Relative Risk*, *Efficacy*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df))
```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

$\pi = \frac{1 - \psi}{2 - \psi} \\ L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162} \\ L(\psi) = {170 \choose 8} (\frac{1 - \psi}{2 - \psi})^8(1 - \frac{1 - \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{2 - \psi - 1 + \psi}{2 - \psi})^{162} = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162} \\ \ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi)) \\ \frac{d}{d\psi}\ell(\psi) = \frac{-8}{1 - \psi} + \frac{8}{2 - \psi} + \frac{162}{2 - \psi} = \frac{170}{2 - \psi} - \frac{8}{1 - \psi} = 0 \\ 170 - 170 \psi - 16 + 8\psi = 0 \\ 162\psi = 154 \\ \hat\psi^{mle}_0 = 0.9506173$  

We know that the likelihood function for $\pi$ will be $L(\pi) = {170 \choose 8} (\pi)^8(1 - \pi)^{162}$, which means the likelihood function of $\psi$ would be $L^*(\psi) = {170 \choose 8}(\frac{1 - \psi}{2 - \psi})^8(\frac{1}{2 - \psi})^{162}$ or $\ell(\psi) = ln({170 \choose 8}) +  8(ln(1 - \psi) - ln(2 - \psi)) + 162(ln(1) - ln(2 - \psi))$. From this, we can see that $\hat\psi^{mle}_0 = 0.9506173$, which would be a global maximum. 

## Bayesian Inference
Detail the Bayesian approach.

```{r, meredith_bayes_ab, echo = F}
ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))
```

From this, we can see that we would have an a value of `r ab[1]` and a b value of `r ab[2]`. This would give us the following models: $$f(T | \pi) \sim Binom(n = 170, \pi) \text{ Data Model}$$ $$g(\pi) = Beta(a = `r ab[1]`, b = `r ab[2]`) \text{ Prior Model}$$ $$h(\pi | t) = Beta(a = `r ab[1]` - t, b = 170 - t + `r ab[2]`)$$

```{r, meredith_bayes_vis}
loglik_binom <- function(psi, n, x){
  log(choose(n, x)) + x * log(1 - psi) - x * log(2 - psi) - (n - x) * log(2 - psi)
}

ggplot() + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")

ggplot() + 
  geom_function(fun = loglik_binom, 
                mapping = aes(color = "Log Likelihood"), 
                args = list(n = 170, x = 8), 
                xlim = c(0, 1), 
                n = 500) +
  geom_vline(xintercept = 0.9506173) + 
  labs(title = "Plot of Log Likelihood", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")

ggplot() + # Icky for some reason, but I think this is what we were supposed to do?
  geom_function(fun = dbeta, # I will look at this again later and see if I can find the error, because something is really wrong
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = loglik_binom, 
                mapping = aes(color = "Log Likelihood"), 
                args = list(n = 170, x = 8), 
                xlim = c(0, 1), 
                n = 500) + 
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")
```

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
df <- data.frame("Vaccine" = c(rep("BNT162b2", times = 8 + 17411), 
                               rep("placebo", times = 162 + 17511)), 
                 "covid" = c(rep("yes", times = 8), 
                             rep("no", times = 17411), 
                             rep("yes", times = 162), 
                             rep("no", times = 17511)))
df %>% tabyl(Vaccine, covid) %>% adorn_percentages()%>% adorn_totals %>% adorn_ns() 

mosaicplot(table(df))

ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))

loglik_binom <- function(psi, n, x){
  log(choose(n, x)) + x * log(1 - psi) - x * log(2 - psi) - (n - x) * log(2 - psi)
}

ggplot() + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")

ggplot() + 
  geom_function(fun = loglik_binom, 
                mapping = aes(color = "Log Likelihood"), 
                args = list(n = 170, x = 8), 
                xlim = c(0, 1), 
                n = 500) +
  geom_vline(xintercept = 0.9506173) + 
  labs(title = "Plot of Log Likelihood", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")

ggplot() + # Icky for some reason, but I think this is what we were supposed to do?
  geom_function(fun = dbeta, # I will look at this again later and see if I can find the error, because something is really wrong
                mapping = aes(color = "Prior"), 
                args = list(shape1 = ab[1], shape2 = ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = dbeta, 
                mapping = aes(color = "Posterior"), 
                args = list(shape1 = ab[1] - 8 + 1, shape2 = 170 - 8 + 1 + ab[2]), 
                xlim = c(0, 1)) + 
  geom_function(fun = loglik_binom, 
                mapping = aes(color = "Log Likelihood"), 
                args = list(n = 170, x = 8), 
                xlim = c(0, 1), 
                n = 500) + 
  labs(title = "Prior vs Posterior vs Likelihood Model of COVID Vaccine", 
       subtitle = "t = 8 observed", 
       x = expression(psi), 
       y = "PDF")
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.
