---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(LearnBayes)
```

# Abstract
Write your abstract here.

# Keywords
*Bayesian Statistics*, *Likelihood Inference*, *Relative Risk*, *Efficacy*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

## Bayesian Inference
Detail the Bayesian approach.

```{r, meredith_bayes_ab, echo = F}
ab <- beta.select(quantile1 = list(p = .01, x = 7/17), 
            quantile2 = list(p = .1, x = 1/2))
```

From this, we can see that we would have an a value of `r ab[1]` and a b value of `r ab[2]`. This would give us the following models: $$f(T | \pi) \sim Binom(n = 170, \pi) \text{ Data Model}$$ $$g(\pi) = Beta(a = `r ab[1]`, b = `r ab[2]`) \text{ Prior Model}$$ $$h(\pi | t) = Beta(a = `r ab[1]` - t, b = 170 - t + `r ab[2]`)$$


# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.
